---
title: "Data Visualization I (ggplot2)"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Data Visualization I (ggplot2)}
  %\VignetteEngine{knitr::rmarkdown}
---

Close down R, say 'No' to saving the workspace.

# <a name="workflow"></a> Workflow

The typical workflow in R is 

1. Start R
1. Choose your working directory
1. Open a script
1. Read in data
1. Check data (interactively)
1. Do analysis

If you are using RStudio, I highly recommend you look into setting up [projects](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects). 

## Choose your working directory 

Start R and then [choose your working directory](intro.html#working_directory).

```{r, eval=FALSE}
setwd(choose.dir())
```

## Start the workshop

Now start the workshop to get your data files and scripts set up. Typically you would choose your working directory to be where you actually have your data and scripts.

```{r}
library(ISDSWorkshop)
workshop(launch_index  = FALSE)
```

## Open the script

Open the `advanced_graphics.R` script.

Notice that the top of the script has all of the packages that you will use.

```{r}
#########################################
# Pretend this is the top of the script #
#########################################
library(plyr)
library(ggplot2)
library(gridExtra)
```


## Read in the data

Typically, you will have a script that reads in the data, converts variables, and then performs statistical analyses.

This time we will use the `mutate()` function from the [plyr](http://cran.r-project.org/web/packages/plyr/index.html) package. This function allows you to create new columns in a `data.frame`.

```{r}
# Read in csv files
GI = read.csv('GI.csv')
icd9df = read.csv("icd9.csv")

# Mutate data.frame
GI = mutate(GI,
            date      = as.Date(date),
            weekC     = cut(date, breaks="weeks"),
            week      = as.numeric(weekC),
            facility  = as.factor(facility),
            icd9class = factor(cut(icd9, 
                            breaks = icd9df$code_cutpoint, 
                            labels = icd9df$classification[-nrow(icd9df)], 
                            right  = TRUE)),
            ageC      = cut(age, 
                            breaks = c(-Inf, 5, 18, 45 ,60, Inf)),
            zip3      = trunc(zipcode/100))
```




## Reading other scripts

If the script to read in the data is extensive, it may be better to separate your scripts into two different files: one for reading in the data and converting variables and another (or more) to perform statistical analyses. Then, in the second file, i.e. the one that does an analysis, you can `source()` the one that just reads in the data, e.g.

```{r, eval=FALSE}
source("read_data.R") 
```


## Check the data

At this point, I typically check the data to make sure it is what I think it is. 
I usually do this in interactive mode, i.e. not in a script.

```{r}
dim(GI)
str(GI)
summary(GI)
```


## Activity

Create a new variable in the GI data set called `weekD` that is a `Date` object with each observation having the Monday of the week for that observation. Check that it is actually a date using the `str()` function.

```{r, echo=FALSE}
# Create weekD variable in GI data set
```


When you have completed the activity, compare your results to the [solutions](advanced_graphics-solution.html#Date).



# <a name="graph_workflow"></a> Graph workflow

Now to construct a graph, the workflow is

1. construct an appropriate data set and
1. construct the graph.


## Construct the data set

Suppose we want to plot the number of weekly GI cases by facility. We need to aggregate the data by week and facility. 

```{r}
GI_wf = ddply(GI, .(week, facility), summarize, count = length(id))
```

In interactive mode, you should verify that this data set is correct, e.g. 

```{r, eval=FALSE}
nrow(GI_wf) # Should have number of weeks times number of facilities rows
ncol(GI_wf) # Should have 3 columns: week, facility, count
dim(GI_wf)
head(GI_wf)
tail(GI_wf)
summary(GI_wf)
summary(GI_wf$facility)
```

## Construct the graph

Now, we would like week on the x-axis and count on the y-axis. 

```{r}
ggplot(GI_wf, aes(x=week, y=count)) + geom_point()
```

But, clearly we need to distinguish the facilities. 


## Try colors and shapes to distinguish facilities

Colors:

```{r}
ggplot(GI_wf, aes(x=week, y=count, color=facility)) + geom_point()
```

Shapes:

```{r}
ggplot(GI_wf, aes(x=week, y=count, shape=facility)) + geom_point()
```

ggplot2 only plots 6 different shapes and provides a warning.


## Activity - weekly GI counts by age category

Construct a data set and then a plot to look at weekly GI cases by age category.

```{r, echo=FALSE}
# Construct a data set aggregated by week and age category

# Construct a plot to look at weekly GI cases by age category.
```

When you have completed the activity, compare your results to the [solutions](advanced_graphics-solution.html#graph).



# <a name="faceting"></a> Faceting

Faceting is a technique to view many small graphs on a single page.

In ggplot, there are two different ways to construct facets: `facet_wrap()` and `facet_grid()`. The former performs the faceting for a single variable and constructs the matrix of plots automatically. The latter performs faceting for two variables putting one vertically and the other horizontally.


## Faceting on a single variable

When there are many levels of a single variable, `facet_wrap()` can often allow comparisons across that variable.

```{r}
ggplot(GI_wf, aes(x=week, y=count)) + geom_point() + facet_wrap(~facility)
```

By default, faceting forces all axes to be exactly the same. If you want the axis to automatically scale for each facet use `scales="free"`.

```{r}
ggplot(GI_wf, aes(x=week, y=count)) + geom_point() + facet_wrap(~facility, scales="free")
```

## Faceting on two variables

When there are two variables (with not very many levels in each), `facet_grid()` can allow relevant comparisons.

Suppose we are interested in weekly GI counts by gender and age category. First, construct the data set

```{r}
GI_sa = ddply(GI, .(week, gender, ageC), summarize, count = length(id))
```

Next, construct the plot using `facet_grid(row~column)` syntax.

```{r}
ggplot(GI_sa, aes(x=week, y=count)) + geom_point() + facet_grid(gender~ageC)
```

## Using faceting and colors/shapes

In addition, we could use faceting together with colors/shapes

```{r}
ggplot(GI_sa, aes(x=week, y=count, shape=gender, color=gender)) + geom_point() + facet_wrap(~ageC)
```


## Activity - weekly GI counts by zip3 and ageC

Construct a plot of weekly GI counts by zip3 and ageC. 

```{r, echo=FALSE}
# Construct a plot of weekly GI counts by zip3 and ageC. 
```

When you have completed the activity, compare your results to the [solutions](advanced_graphics-solution.html#facet).




# <a name="subsetting"></a> Subsetting

Often our data set is large and we would like to [subset the data](intro.html#subsetting) to focus on a particular group or time frame. You may want to subset by a

- categorical, 
- continuous (numeric), or 
- `Date` variable.

## Subsetting by a categorical variable

For a categorical variable, you want to subset by 

- a single category
- a set of categories
- not in a single category
- not in a set of categories

### Subsetting by a single category

Suppose we are only interested in counts specifically for infectious and parasitic diseases (IPD).

```{r}
IPD = subset(GI, icd9class == "infectious and parasitic disease")
IPD_w = ddply(IPD, .(week), summarize, count=length(id))
ggplot(IPD_w, aes(x=week, y=count)) + geom_point()
```


### Subsetting by a set of categories

Suppose we are interested in looking at both IPD and ill-defined conditions

```{r}
conditions = levels(GI$icd9class)
conditions
```

Say we want levels 1 and 3. 

```{r}
IPDp = subset(GI, icd9class %in% levels(icd9class)[c(1,3)])
summary(IPDp$icd9class)
```

Now construct the graph

```{r}
IPDp_w = ddply(IPDp, .(week, icd9class), summarize, count = length(id))
ggplot(IPDp_w, aes(x=week, y=count)) + geom_point() + facet_wrap(~icd9class, scales="free")
```

### Subsetting by not in a single category

```{r}
nIPD = subset(GI, icd9class != "infectious and parasitic disease")
nIPD_w = ddply(nIPD, .(week, icd9class), summarize, count = length(id))
ggplot(nIPD_w, aes(x=week, y=count)) + geom_point() + facet_wrap(~icd9class, scales="free")
```

### Subsetting by not in a set of categories

```{r}
nIPDp = subset(GI, !(icd9class %in% levels(icd9class)[c(1,3)]))
nIPDp_w = ddply(nIPDp, .(week, icd9class), summarize, count = length(id))
ggplot(nIPDp_w, aes(x=week, y=count)) + geom_point() + facet_wrap(~icd9class, scales="free")
```


## Subsetting by a continuous (numeric) variable

We may want to subset a continuous (numeric) variable by

- Larger than (or equal to) a number
- Small than (or equal to) a number
- In a range of numbers (with our without endpoints)

### Subsetting larger than a number

```{r}
Age60p = subset(GI, age > 60)
Age60p_w = ddply(Age60p, .(week), summarize, count = length(id))
ggplot(Age60p_w, aes(x=week, y=count)) + geom_point()
```

### Subsetting less than or equal to a number 

```{r}
Age_lte30 = subset(GI, age <= 30)
Age_lte30_w = ddply(Age_lte30, .(week), summarize, count = length(id))
ggplot(Age_lte30_w, aes(x=week, y=count)) + geom_point()
```

### Subsetting within a range

```{r}
Age30to60 = subset(GI, age > 30 & age <= 60) # This includes those who are 60, but not those who are 30
Age30to60_w = ddply(Age30to60, .(week), summarize, count = length(id))
ggplot(Age30to60_w, aes(x=week, y=count)) + geom_point()
```

## Subsetting a date

Subsetting a date works similar to subsetting a continuous (numeric) variable once you convert a comparison value to a `Date` using the `as.Date()` function.

Let's create a `Date` vector containing the two values we will use for subsetting.

```{r}
two_dates = as.Date(c("2007-01-01", "2007-12-31"))
```

### Subsetting dates

```{r}
early = subset(GI, date <  two_dates[1])
late  = subset(GI, date >= two_dates[2])
mid   = subset(GI, date >= two_dates[1] & date < two_dates[2])
```

Now make plots

```{r}
early_w = ddply(early, .(week), summarize, count = length(id))
late_w  = ddply(late,  .(week), summarize, count = length(id))
mid_w   = ddply(mid,   .(week), summarize, count = length(id))

g1 = ggplot(early_w, aes(x=week, y=count)) + geom_point()
g2 = ggplot(late_w,  aes(x=week, y=count)) + geom_point()
g3 = ggplot(mid_w,   aes(x=week, y=count)) + geom_point()
```

To arrange multiple plots, use the `grid.arrange` function in the `gridExtra` package.

```{r, eval=FALSE}
library(gridExtra)
```

```{r}
grid.arrange(g1,g3,g2)
```


## Activity - subsetted graphs

Construct a plot for those in zipcode 206xx between Jan 1, 2008 and Dec 31, 2008. 

```{r, echo=FALSE}
# Subset the data to zipcode 206xx between Jan 1, 2008 and Dec 31, 2008

# Aggregate the date for each week in this time frame

# Construct the plot of weekly GI counts in zipcode 206xx.
```

When you have completed the activity, compare your results to the [solutions](advanced_graphics-solution.html#subsetting).



# <a name="professional"></a> Making professional looking graphics

To make the graphics we are producing look professional, we will need to [customize](graphics.html#customizing) the graph according to the final production platform, e.g. website, word document, pdf document, etc. This will be highly specific to your use. Nonetheless, likely you will use the following `ggplot2` functions:

- `scale_color_manual`
- `labs`
- `theme`

But there are many others. One thing you will notice is that we can keep updating the graph by updating the R object that holds the graph. 

## Base graphic

Let's start with the following graph

```{r}
GI$weekD = as.Date(GI$weekC) 
GI_sum = ddply(GI, .(weekD, gender, ageC), summarize, count = length(id))
ggplot(GI_sum, aes(x=weekD, y=count, color=gender)) + geom_point(size=3) + facet_grid(ageC~., scales="free_y")
```

## Change age category labels

The easiest way to change some of the labels on the graph is to change the underlying data. Here we change the levels associated with the `ageC` variable. 

```{r}
levels(GI_sum$ageC)
levels(GI_sum$ageC) = paste("Age:", c("<5","5-18","18-45","45-60",">60"))
table(GI_sum$ageC)
```

```{r}
g = ggplot(GI_sum, aes(x=weekD, y=count, color=gender)) + geom_point(size=3) + facet_grid(ageC~., scales="free_y")
g
```

`g` is the R object that contains the graph


## Change colors 

Intuitive colors can make a graph easier to read. 

```{r}
g = g + scale_color_manual(values=c("hotpink","blue"), name='Sex')
g
```


## Change title and axis labels

Use proper names and capitalization in title and axis labels.

```{r}
g = g + labs(title="Weekly GI cases", x="Year", y="Weekly count") 
g
```

## Try alternate themes

Try alternate themes:

```{r}
g = g + theme_bw()
g
```

To see a list of possibilities:

```{r, eval=FALSE}
?theme_bw
```


## Adjust font sizes

Depending on the final platform, font sizes may need to be adjusted.

```{r}
g = g + theme(title=element_text(size=rel(2)),
              text = element_text(size=16),
              legend.background = element_rect(fill="white", size=.5, color="black"))
g
```


## <a name="exporting"></a> Exporting graphs

To export the graphs, use the `ggsave()` function which saves the last plot that you displayed. As a default, it uses the size of the current graphics device, but you will likely want to modify this. 

```{r}
g
ggsave("plot.pdf")
```

If you look in your current working directory, you will see the plot.pdf file. 

```{r}
ggsave("plot.pdf", width=14, height=8)
```
