---
title: "Using R as a calculator"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using R as a calculator}
  %\VignetteEngine{knitr::rmarkdown}
---

# Using R as a calculator

## Basic calculator operations

All basic calculator operations can be performed in R. 

```{r}
1+2
1-2
1/2
1*2
```



## Advanced calculator operations

Many advanced calculator operations are also available.

```{r}
(1+3)*2+100^2 # standard order of operations
sin(2*pi)
sqrt(4)
10^2
log(10)
log(10,base=10)
```



## Using variables

A real advantage to using R rather than a calculator (or calculator app) is the ability to store quantities using variables.

```{r}
a = 1
b = 2
a+b
a-b
a/b
a*b
```



## Using informative variable names

While using variables alone is useful, it is much more useful to use informative variables names.

```{r}
population = 1000
number_infected = 200
deaths = 3

death_rate = deaths / number_infected
attack_rate = number_infected / population

death_rate
attack_rate
```



## Activity - Bayes' Rule

Suppose an individual tests positive for a disease, what is the probability the individual has the disease? Let 

- $D$ indicates the individual has the disease
- $N$ means the individual does not have the disease
- $+$ indicates a positive test result
- $-$ indicates a negative test 

The above probability can be calculated using [Bayes' Rule](http://en.wikipedia.org/wiki/Bayes%27_rule):

\[ P(D|+) = \frac{P(+|D)P(D)}{P(+|D)P(D)+P(+|N)P(N)} = \frac{P(+|D)P(D)}{P(+|D)P(D)+(1-P(-|N))\times(1-P(D))} \]

where

- $P(+|D)$ is the [sensitivity](http://en.wikipedia.org/wiki/Sensitivity_and_specificity) of the test
- $P(-|N)$ is the [specificity](http://en.wikipedia.org/wiki/Sensitivity_and_specificity) of the test
- $P(D)$ is the [prevalence](http://en.wikipedia.org/wiki/Prevalence) of the disease

Calculate the probability the individual has the disease if the test is positive when

- the specificity of the test is 0.99, 
- the sensistivity of the test is 0.95, and
- the prevalence of the disease is 0.001.


When you have completed the activity, compare your results to the [solutions](calculator-solution.html).



