% Using R as a fancy calculator
% Jarad Niemi
% 2014/06/23
<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Using R as a fancy calculator}
-->



```{r setup, include=FALSE}
library('knitr')
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center")
```


# Basic calculator functions in R



```{r}
1+1
1-1
4/2
4*2
```


# Advanced calculator functions in R


```{r}
sqrt(4)
abs(-4)
cos(2*pi)
```


# Sometimes the functions aren't what you are expecting

```{r}
log(10)
log10(10)
log(10,base=10)
log(10,base=exp(1))
```

# Using variables


```{r}
a = 2
b = 3
a*b
b/a
a+b
a-b
```


# Using informative names


```{r}
number_of_cases = 10

population = 1000

attack_rate = number_of_cases / population

attack_rate

number_of_cases = 20

number_of_cases / population
```




# Bayes' rule

In diagnostic testing, we calculate the probability of a disease after a positive test using the Bayes' rule: 

\[] P(D|+) = \frac{P(+|D)P(D)}{P(+|D)P(D)+P(+|\sim D)P(\sim D)} \]

where 

- P(+|D) is the sensitivity of the test
- P(-|~D) is the specificity of the test
- P(D) is the prevalence of the disease


