---
title: "Calculator"
output:
  html_document:
    toc: true
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Using R as a calculator}
-->


# Basic calculator operations

All basic calculator operations can be performed in R. 

```{r}
1+2
1-2
1/2
1*2
```



# Advanced calculator operations

Many advanced calculator operations are also available.

```{r}
(1+3)*2+100^2 # standard order of operations
sin(2*pi)
sqrt(4)
10^2
log(10)
log(10,base=10)
```



# Using variables

A real advantage to using R rather than a calculator (or calculator app) is the ability to store quantities using variables.

```{r}
a = 1
b = 2
a+b
a-b
a/b
a*b
```



# Using informative variable names

While using variables alone is useful, it is much more useful to use informative variables names.

```{r}
population = 1000
number_infected = 200
deaths = 3

death_rate = deaths / number_infected
attack_rate = number_infected / population

death_rate
attack_rate
```



# Activity

Practice using the calculator functionality of R with this [activity](calculator-activity.html).



