---
title: "Graphics in R (ggplot2)"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Graphics in R (ggplot2)}
  %\VignetteEngine{knitr::rmarkdown}
---


Before we get back into graphics, it is important to understand some of the fundamentals behind what R is doing. Please open the `graphics.R` script in your working directory. 

# <a name="types"></a> Data types

Objects in R can be broadly classified as according to their dimensions:

- scalar
- vector
- matrix
- array (higher dimensionl matrix)

and according to the type of variable they contain:

- numeric
- character
- logical
- factor


## Scalars

Scalars only have a single value assigned to the object in R. 

```{r}
a = 3.14159265 
b = "ISDS Workshop" 
c = TRUE
```

Print the object

```{r}
a
b
c
```


## Vectors

The `c()` function creates a vector in R

```{r}
a = c(1,2,-5,3.6)
b = c("ISDS","Workshop")
c = c(TRUE, FALSE, TRUE)
```

To determine the length of a vector in R use `length()`

```{r}
length(a)
length(b)
length(c)
```


### Vector construction

Create a numeric vector that is a sequence
```{r}
1:10
5:-2
seq(2,23,by=3)
```

Another useful function to create vectors is `rep()`

```{r}
rep(1:4, times = 2)
rep(1:4, each  = 2)
rep(1:4, each  = 2, times = 2)
```


### Accessing vector elements 

Elements of a vector can be access using the index 

```{r}
a = c("one","two","three","four","five")
a[2:4]
a[c(3,5)]
a[rep(3,4)]
```

or by using a logical vector

```{r}
a[c(TRUE, TRUE, FALSE, FALSE, FALSE)]
```

You can also remove elements

```{r}
a[-1]
```


## Matrices

Matrices can be constructed using `cbind()`, `rbind()`, and `matrix()`:

```{r}
m1 = cbind(c(1,2), c(3,4))       # Row bind
m2 = rbind(c(1,3), c(2,4))       # Column bind
all.equal(m1, m2)
m3 = matrix(1:4, nrow=2, ncol=2)
all.equal(m1,m3)
m4 = matrix(1:4, nrow=2, ncol=2, byrow=TRUE)
all.equal(m3, m4)
m3
m4
```


### Accessing matrix elements

Elements of a matrix can be accessed using their indices `[row,column]`

```{r}
m = matrix(1:12, nrow=3, ncol=4)
m
m[2,3]
```

Multiple elements can be accessed at once

```{r}
m[1:2,3:4]
```

If no row (column) index is provided, then the whole row (column) is accessed.

```{r}
m[1:2,]
```

Like vectors, you can eliminate rows (or columns)

```{r}
m[-c(3,4),]
```

Be careful not to forget the comma

```{r}
m[1:2]
```


## Cannot mix types 

You cannot mix types within a vector, matrix, or array

```{r}
c(1,"a")
```

The number 1 is in quotes indicating that R is treating it as a character rather than a numeric.

```{r}
c(TRUE, 1, FALSE)
```

The logicals are converted to numeric (0 for FALSE and 1 for TRUE).

```{r}
c(TRUE, 1, "a")
```

## Activity

Reconstruct the following matrix using the matrix function, then 

1. Print the element in the 3rd-row and 4th column
1. Print the 2nd column
1. Print all but the 4th row

```{r}
m = rbind(c(1, 12, 8, 6),
          c(4, 10, 2, 9),
          c(11, 3, 5, 7))
m
```

```{r, echo=FALSE}
# Reconstruct the matrix 

# Print the element in the 3rd-row and 4th column

# Print the 2nd column

# Print all but the 3rd row
```


When you have completed the activity, compare your results to the [solutions](graphics-solution.html#matrix).




# <a name="frames"></a> Data frames

A `data.frame` is a special type of matrix that allows different data types in different columns.

We have already seen a data frame with our GI data set. Let's read this data in again and take a look.

```{r}
GI = read.csv("GI.csv")
dim(GI)
```

If the above didn't work, try

```{r, eval=FALSE}
workshop()
GI = read.csv("GI.csv")
dim(GI)
```


## Access `data.frame` elements

Data frames can be accessed just like matrices

```{r}
GI[1:2, 3:4]
```

but can also be accessed by column names

```{r}
GI[1:2, c("facility","icd9","gender")]
```


## Different data types in different columns

The function `str()` allows you to see the structure of any object in R

```{r}
str(GI)
```



## <a name="factor"></a> Factor

A factor is a data type that represents a categorical variable in a `data.frame`. 

The default is for any character vector to be converted to a factor when read into a `data.frame` in R. 

Internally, R codes a factor as an integer and then keeps a table that contains the conversion from that integer into the actual value of the factor.

```{r}
nlevels(GI$gender)
levels(GI$gender)
GI$gender[1:3]
as.numeric(GI$gender[1:3])
```


### Converting a numeric variable into a factor

When categorical variables are encoded as numeric variables in the data set, R reads them in as numeric. To convert them to a factor use `as.factor()`. 

```{r}
GI$facility = as.factor(GI$facility)
table(GI$facility)
```


### Creating your own factor

Use the `cut()` function to create a factor from a continuous variable.

```{r}
GI$ageC = cut(GI$age, c(-Inf, 5, 18, 45 ,60, Inf))
table(GI$ageC)
```

This created a new variable in the GI `data.frame` called `ageC`. `Inf` represents infinity.



## <a name="dates"></a> Dates

In order to use dates properly, they need to be converted into type `Date`. 

```{r}
GI$date = as.Date(GI$date)
str(GI$date)
```

`as.Date()` will attempt to read dates as "%Y-%m-%d" then "%Y/%m/%d". If neither work, it will give an error. 

```{r, eval=FALSE}
?as.Date
```

You can specify other date patterns, e.g. 

```{r, eval=FALSE}
as.Date("12/09/14", format="%m/%d/%y")
```


## Activity 

Create a new variable in the GI data set called `icd9code` that cuts icd9 at 0, 140, 240, 280, 290, 320, 360, 390, 460, 520, 580, 630, 680, 710, 740, 760, 780, 800, and Inf. Find the icd9code that is the most numerous in the GI data set.

```{r, echo=FALSE}
# Create icd9code

# Find the icd9code that is most numerous
```



When you have completed the activity, compare your results to the [solutions](graphics-solution.html#factor).




# <a name="reshaping"></a> Reshaping data frames in R

There are two general representations of tabular data.

Wide:
```{r, echo=FALSE}
d = data.frame(week=1:3, GI=c(246,195,212), ILI=c(948, 1020, 1024))
d
```
which is a succinct representation of the data

Long:
```{r, echo=FALSE}
library(reshape2)
melt(d, id.vars="week", variable.name="syndrome", value.name="count")
```
which is the form most statistical software wants, i.e. there is only one column for the response (count).


## Wide to long

The `reshape2` package provides functions to convert between the two representations. First, we need to load the package

```{r}
library(reshape2)
```

Create the wide `data.frame`:
```{r}
d = data.frame(week=1:3, GI=c(246,195,212), ILI=c(948, 1020, 1024))
```

To turn the `data.frame` into long format, use `melt()`

```{r}
m = melt(d, id.vars    = "week",     # The variables you want to remain on the columns
         variable.name = "syndrome", # The name for the variable column
         value.name    = "count")    # The name for the response column
m
```


## Long to wide

If we want to convert back, use `dcast()`

```{r}
dcast(m, week ~ syndrome) # Notice that we do not use the count (value) column at all
```


# <a name="aggregating"></a> Aggregating data frames in R

The GI data set that we have is already in long format and each row is an individual. 
We may want to aggregate this information. To do so, we will use the `ddply()` function in the `plyr` package.

```{r}
library(plyr)
```

For example, perhaps we wanted to know the total number of GI or ILI cases across the 3 weeks:

```{r}
ddply(m,                  # We need to use the melted version of the data set
      .(syndrome),        # Do the following for each syndrome
      summarize,          # Summarize 
      total = sum(count)) # Calculate `total` which is the sum of count for each syndrome
```


## Aggregating the GI data set

Let's aggregate the GI data set by week, gender, and age category.

First, we need to create weeks

```{r}
GI$date = as.Date(GI$date) # Make sure the dates are actually dates
GI$week = cut(GI$date, 
              breaks="weeks", 
              start.on.monday=TRUE) 
```

Now we can summarize

```{r}
GI_count = ddply(GI, 
                 .(week, gender, ageC),
                 summarize,
                 total = length(id))    # Count the number of lines in each week-gender-ageC combination
GI_count
```


## Activity

Aggregate the GI data set by gender, ageC, and icd9code (the ones created in the last activity).


```{r, echo=FALSE}
# Aggregate our GI data set by gender, ageC, and icd9code (the ones created in the last activity).
```


When you have completed the activity, compare your results to the [solutions](graphics-solution.html#aggregate).





# <a name="ggplot2"></a> Basics of `ggplot2`

Once you have your data in long format, you can use the `ggplot2` package for graphics.

Load the `ggplot2` package
```{r}
library(ggplot2)
```



## Histogram

A basic histogram in ggplot

```{r, eval=FALSE, echo=FALSE}
qplot(data=GI, x=age, geom="histogram", binwidth=1)
```

```{r}
ggplot(GI, aes(x = age)) + geom_histogram(binwidth=1)
```


## Boxplots

A basic boxplot

```{r}
ggplot(GI, aes(x = 1, y = age)) + geom_boxplot()
```


## Multiple boxplots

```{r}
ggplot(GI, aes(x = facility, y = age)) + geom_boxplot()
```



## Scatterplots

```{r}
ggplot(GI, aes(x=date, y=age)) + geom_point()
```

## Bar charts

With ggplot, there is no need to count first.

```{r}
ggplot(GI, aes(x=facility)) + geom_bar()
```



## Activity


Construct a histogram and boxplot for age at facility 37 using ggplot2.

```{r, echo=FALSE}
# Construct a histogram for age at facility 37.

# Construct a boxplot for age at facility 37. 
```

Construct a bar chart for the zipcode at facility 37 using ggplot2

```{r, echo=FALSE}
# Construct a bar chart for the first three digits of zipcode at facility 37.
```


When you have completed the activity, compare your results to the [solutions](graphics-solution.html#ggplot2).




# <a name="customizing"></a> Customizing ggplot2 plots

There are many ways to customize the appearance of ggplot2 plots:

- Colors
- Labels
- Titles
- Characters
- Line types
- Themes



## Colors

```{r}
ggplot(GI, aes(x = age)) + geom_histogram(binwidth=1, color='red',   fill='blue')
ggplot(GI, aes(x = age)) + geom_histogram(binwidth=1, color='white', fill='orange')
```

```{r}
ggplot(GI, aes(x=date, y=age)) + geom_point(color='purple')
```


## Labels

```{r}
ggplot(GI, aes(x = facility, y = age)) + 
  geom_boxplot() + 
  labs(x='Facility ID', y='Age (in years)', title='Age by Facility ID')
```


## Characters

```{r}
ggplot(GI, aes(x=date, y=age)) + geom_point(shape=2, color='red')
```

ggplot2 uses the same shape codes as base graphics, see

```{r, eval=FALSE}
?points
```


## Line types

```{r}
g = ggplot(ddply(GI, .(week), summarize, count=length(id)),
       aes(x=as.numeric(week), y=count)) +
  labs(x='Week #', y='Weekly count')

g + geom_line()
g + geom_line(size=2, color='firebrick', linetype=2)
```

## Themes

```{r}
g = g+geom_line(size=1, color='firebrick')
g + theme_bw()
```

For other themes, see

```{r, eval=FALSE}
?theme
?theme_bw
```




# <a name="help"></a> Getting help on ggplot2

Although the general R help can still be used, e.g. 

```{r, eval=FALSE}
?ggplot
?geom_point
```

It is much more helpful to google for an answer

    geom_point 
    ggplot line colors

The top hits will all have the code along with what the code produces.

## Helpful sites

These sites all provide code. The first two also provide the plots that are produced.

- <http://docs.ggplot2.org/current/>
- <http://www.cookbook-r.com/Graphs/>
- <http://stackoverflow.com/questions/tagged/ggplot2>



## Activity

Play around with ggplot2 to see what kind of changes you can make.

