---
title: "Graphics in R (ggplot2)"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Graphics in R (ggplot2)}
  %\VignetteEngine{knitr::rmarkdown}
---


Before we get back into graphics, it is important to understand some of the fundamentals behind what R is doing. Please open the `graphics.R` script in your working directory. 

# Data types

Objects in R can be broadly classified as according to their dimensions:

- scalar
- vector
- matrix
- array (higher dimensionl matrix)

and according to the type of variable they contain:

- numeric
- character
- logical
- factor


## Scalars

Scalars only have a single value assigned to the object in R. 

```{r}
a = 3.14159265 
b = "ISDS Workshop" 
c = TRUE
```

Print the object

```{r}
a
b
c
```


## Vectors

The `c()` function creates a vector in R

```{r}
a = c(1,2,-5,3.6)
b = c("ISDS","Workshop")
c = c(TRUE, FALSE, TRUE)
```

To determine the length of a vector in R use `length()`

```{r}
length(a)
length(b)
length(c)
```


### Vector construction

Create a numeric vector that is a sequence
```{r}
1:10
5:-2
seq(2,23,by=3)
```

Another useful function to create vectors is `rep()`

```{r}
rep(1:4, times = 2)
rep(1:4, each  = 2)
rep(1:4, each  = 2, times = 2)
```


### Accessing vector elements 

Elements of a vector can be access using the index 

```{r}
a = c("one","two","three","four","five")
a[2:4]
a[c(3,5)]
a[rep(3,4)]
```

or by using a logical vector

```{r}
a[c(TRUE, TRUE, FALSE, FALSE, FALSE)]
```

You can also remove elements

```{r}
a[-1]
```


## Matrices

Matrices can be constructed using `cbind()`, `rbind()`, and `matrix()`:

```{r}
m1 = cbind(c(1,2), c(3,4))       # Row bind
m2 = rbind(c(1,3), c(2,4))       # Column bind
all.equal(m1, m2)
m3 = matrix(1:4, nrow=2, ncol=2)
all.equal(m1,m3)
m4 = matrix(1:4, nrow=2, ncol=2, byrow=TRUE)
all.equal(m3, m4)
m3
m4
```


### Accessing matrix elements

Elements of a matrix can be accessed using their indices `[row,column]`

```{r}
m = matrix(1:12, nrow=3, ncol=4)
m
m[2,3]
```

Multiple elements can be accessed at once

```{r}
m[1:2,3:4]
```

If no row (column) index is provided, then the whole row (column) is accessed.

```{r}
m[1:2,]
```

Like vectors, you can eliminate rows (or columns)

```{r}
m[-c(3,4),]
```

Be careful not to forget the comma

```{r}
m[1:2]
```


## Cannot mix types 

You cannot mix types within a vector, matrix, or array

```{r}
c(1,"a")
```

The number 1 is in quotes indicating that R is treating it as a character rather than a numeric.

```{r}
c(TRUE, 1, FALSE)
```

The logicals are converted to numeric (0 for FALSE and 1 for TRUE).

```{r}
c(TRUE, 1, "a")
```

## Activity

Reconstruct the following matrix using the matrix function, then 

1. Print the element in the 3rd-row and 4th column
1. Print the 2nd column
1. Print all but the 4th row

```{r}
m = rbind(c(1, 12, 8, 6),
          c(4, 10, 2, 9),
          c(11, 3, 5, 7))
m
```

```{r, echo=FALSE}
# Reconstruct the matrix 

# Print the element in the 3rd-row and 4th column

# Print the 2nd column

# Print all but the 3rd row
```





# Data frames

A `data.frame` is a special type of matrix that allows different data types in different columns.

We have already seen a data frame with our GI data set. Let's read this data in again and take a look.

```{r}
GI = read.csv("GI.csv")
dim(GI)
```

If the above didn't work, try

```{r, eval=FALSE}
workshop()
GI = read.csv("GI.csv")
dim(GI)
```


## Access `data.frame` elements

Data frames can be accessed just like matrices

```{r}
GI[1:2, 3:4]
```

but can also be accessed by column names

```{r}
GI[1:2, c("facility","icd9","gender")]
```


## Different data types in different columns

The function `str()` allows you to see the structure of any object in R

```{r}
str(GI)
```



## Factor

A factor is a data type that represents a categorical variable in a `data.frame`. 

The default is for any character vector to be converted to a factor when read into a `data.frame` in R. 

Internally, R codes a factor as an integer and then keeps a table that contains the conversion from that integer into the actual value of the factor.

```{r}
nlevels(GI$gender)
levels(GI$gender)
GI$gender[1:3]
as.numeric(GI$gender[1:3])
```


## Dates

In order to use dates properly, they need to be converted into type `Date`. 

```{r}
GI$date = as.Date(GI$date)
str(GI$date)
```

`as.Date()` will attempt to read dates as "%Y-%m-%d" then "%Y/%m/%d". If neither work, it will give an error. (see `?as.Date`)

You can specify other date patterns, e.g. 

```{r, eval=FALSE}
as.Date("12/09/14", format="%m/%d/%y")
```


## Activity 


