---
title: "Introduction to R - Activity Solutions"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to R - Activity Solutions}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo=FALSE}
library(ISDSWorkshop)
workshop(launch_index=FALSE)
```


# <a name="introduction"></a> Introduction

```{r}
x = 1:10
y = rep(c(1,2), each=5)
m = lm(y~x)
s = summary(m)
```

Now, look at the result of each line

```{r}
x
y
m
s
s$r.squared
```

For those who are familiar with linear regression, this may look familiar. 



# <a name="calculator"></a> Calculator

Calculate the probability the individual has the disease if the test is positive when

- the specificity of the test is 0.99, 
- the sensistivity of the test is 0.95, and
- the prevalence of the disease is 0.001.

```{r}
specificity = 0.99
sensitivity = 0.95
prevalence = 0.001
probability = (specificity*prevalence) / (specificity*prevalence + (1-sensitivity)*(1-prevalence))
probability
```

Yes, it is only about 2%!


# <a name="read"></a> Read csv file

Read in the `fluTrends.csv` file.

```{r}
# Read in the csv file
fluTrends = read.csv('fluTrends.csv')
names(fluTrends)

# To maintain pretty column names, use 
fluTrends = read.csv('fluTrends.csv', check.names=FALSE)
names(fluTrends)
# unfortunately these names won't work with the 
# fluTrends$colname syntax
```


# <a name="descriptive"></a> Descriptive statistics

```{r, echo=FALSE}
GI = read.csv("GI.csv")
```

```{r}
# Min, max, mean, and median age for zipcode 20032.
GI_20032 = subset(GI, zipcode==20032)
min(   GI_20032$age)
max(   GI_20032$age)
mean(  GI_20032$age)
median(GI_20032$age)
```

Alternatively

```{r}
summary(GI_20032$age)
```




# <a name="graphical"></a> Graphical statistics

Construct a histogram and boxplot for age at facility 37.

```{r}
# Construct a histogram and boxplot for age at facility 37.
GI_37 = subset(GI, facility==37)
hist(GI_37$age)

# Construct a boxplot for age at facility 37.
boxplot(GI_37$age)
```

Construct a bar chart for the zipcode at facility 37.

```{r}
# Construct a bar chart for the zipcode at facility 37.
barplot(table(GI_37$zipcode))
```

Perhaps this plot isn't so useful. Maybe it would be better to just use the first 3 zipcode digits

```{r}
# Construct a bar chart for the first three digits of zipcode at facility 37.
barplot(table(trunc(GI_37$zipcode/100)))
```
