---
title: "Introduction to R"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to R}
  %\VignetteEngine{knitr::rmarkdown}
---


# Introduction to R

For an extremely detailed introduction, please see

```{r, eval=FALSE}
help.start()
```

In this documentation, the above command will be executed at the command prompt, see below. 


# R interface

In contrast to many other statistical software packages that use a point-and-click interface, e.g. SPSS, JMP, Stata, etc, R has a command-line interface. The command line has a command prompt, e.g. `>`, see below.

```{r, eval=FALSE}
>
```

This means, that you will be entering commands on this command line and hitting enter to execute them, e.g. 

```{r, eval=FALSE}
help()
```

Use the `up arrow` to recover past commands.

```{r, eval=FALSE}
hepl()
help() # Use up arrow and fix
```


# R GUI (or [RStudio](http://www.rstudio.com/))

Most likely, you are using a graphical user interface (GUI) and therefore, in addition, to the command line, you also have a windowed version of R with some point-and-click options, e.g. File, Edit, and Help.

In particular, there is an editor to create a new R script. So rather than entering commands on the command line, you will write commands in the script and then send those commands to the command line using `Ctrl-R` (PC) or `Command-Enter` (Mac).

```{r}
a = 1 
b = 2
a+b
```

Multiple lines can be run in sequence by selecting them and then using `Ctrl-R` (PC) or `Command-Enter` (Mac).




## Activity

One of the most effective ways to use this documentation is to cut-and-paste the commands into a script and then execute them. 

Cut-and-paste the following commands into a **new script** and then run those commands directly from the script using `Ctrl-R` (PC) or `Command-Enter` (Mac).

```{r}
x = 1:10
y = rep(c(1,2), each=5)
m = lm(y~x)
s = summary(m)
```

Now, look at the result of each line

```{r, eval=FALSE}
x
y
m
s
s$r.squared
```

When you have completed the activity, compare your results to the [solutions](intro-solution.html).

